Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-11-17 response_to_referees.tex

%\documentclass[12pt]{article}
\documentclass[aps,rmp,onecolumn]{revtex4-1}
\usepackage{color}
\usepackage{enumitem}
%\usepackage{natbib}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\comment}[1]{{\color{red}#1}}
\definecolor{response}{rgb}{0.1, 0.1, 0.1}
\definecolor{drab}{rgb}{0.59, 0.44, 0.09}
\definecolor{celestialblue}{rgb}{0.29, 0.59, 0.82}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\definecolor{deepsaffron}{rgb}{1.0, 0.6, 0.2}
\newcommand{\Richard}[1]{{\color{drab}Richard: #1}}
\newcommand{\refa}[1]{\textbf{R1:} #1}
\newcommand{\refb}[1]{\textbf{R2:} #1}
\newcommand{\editor}[1]{\textbf{Editor:} #1}
%\newcommand{\criticism}[1]{\textbf{Criticism:} #1}
\newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}}

\begin{document}
\section*{Response to reviewers}

\subsection*{Editor}

\response{
Thank you very much for handling our manuscript and providing these two constructive reviews that have helped us to improve the paper. 
}

\editor{The comments and recommendations from two expert reviewers are now available for your manuscript. These reviewers judged the reported discoveries to be of medium significance, and the potential scientific impact of your work to be medium/high. They found that the manuscript needs improvement in text and additional data analysis. Editors generally agree with their concerns and recommendations, which led to a designation of medium/high priority.

Note that many manuscripts receiving medium priority based on reviewer comments are not accepted by the Board of Editors. If appropriate, the board may invite a resubmission following the rejection, which is intended to enable you to improve the manuscript towards receiving a high or top priority. The authors should pay close attention to the detailed review comments and address each comment with significant improvements.}

\response{}

\subsection*{Reviewer: 1}

\refa{The manuscript by Barrat-Charlaix et al. discusses the problem of
predictability of mutations in seasonal influenza viruses. It presents
data analysis of frequency trajectories in H1N1 and H3N2 lineages and
simulations of different selection scenarios.
It is an interesting revisit to predictions in influenza, however I
have some major concerns about the analysis and formulation of the
conclusions that should be addressed:}

\refa{1. The prediction problem is defined for each mutation: based on the
tracked frequency trajectory, can the future (fixation, loss, or
polymorphism) of the mutation be predicted. Such formulation has been
previously proposed by Illingworth and Mustonen, (eg. Genetics 2011,
Plos Pathogens 2012). By averaging over all amino-acid substitutions,
the authors show that the fates of mutations are not determined by the
value of the starting frequency.}

\refa{1.1. Mutations in influenza are highly nested, with a substantial
hitch-hiking, and no attempt is made to disentangle such dependencies when counting the mutations.}

\refa{1.2 Despite the more general formulation in the beginning, this
approach makes use only of the last time point in the trajectory,
rather than the full trace (contrary to the work of Illingworth\&
Mustonen). I find the observed neutral-like statistics not surprising
for such a limited data input, which doesn't capture past frequency dynamics. Therefore, these conclusions should be
revisited and benchmarked against the more general implementation of Illingworth\&Mustonen-like approach.}

\refa{2. The authors examined the predictive power of one predictive
method, the LBI. However, they did not do a systematic comparison of
the different methods that they cite (Morris et al, 2018), which differ in prediction
targets and methods. Therefore, the general conclusions about
predictability, eg. on page 8, in line 47, and on page 9, line 48 are
too sweeping and should be made precisely for those methods looked at
in detail.}

\refa{3. Going through the previous and cited literature, I think the authors should cite some of these works in a more careful way. Specifically, the very related work by Illingworth and Mustonen is not mentioned. The distribution $P_delta_t(f|f_0)$ has been at the core of
the method of Strelkowa and Lassig, 2012 (which is cited, but at other
parts of the text), and the same paper also uses a similar simulation model, which should be acknowledged at the appropriate points of the text.}

\refa{It would help if the plots with red-green-blue lines were also distinguished by differing markers.}

\subsection*{Reviewer: 2}

\refb{The paper does a retrospective study of amino acid substitutions in seasonal Influenza to determine what properties of these substitutions could help predict their fate in the future. The authors find that future frequency trajectories are surprisingly unpredictable. Even predicting which mutations fix in the population is hard. The authors find that the current frequency of a mutation is the best predictor for the probability of fixation, which would be expected under neutrality but not in a model with selection. 
I appreciate this study, I think it will be of interest to many readers and it is generally well done and fairly easy to follow. 
}

\refb{1. The authors focus on one feature at a time (frequency, epitope status, LBI …). It is interesting to see that each of these is not very predictive of future frequency / prob of fixation. However, I think the obvious next step is to see whether a combination of many features could do a better job of predicting. I am not sure why the authors don’t try to fit a model that takes into account all information they have about sites (say, type of AA change, location in the gene, current frequency etc) and see if a ML model is able to make predictions. }

\refb{2. Fig 2A and 2B look quite different to me. In the text it appears to me as if these are very similar to the authors. What are the characteristics of the AAs that fix in H1N1? }

\refb{3. Question: what do these results mean for vaccines? How to decide which strains to use for vaccines? This may be known to those who work on Influenza, but for a relative outsider it is not clear. }

\refb{4. Question: at what point is it expected that the vaccine itself will influence frequency trajectories? See Wen et al Biorxiv 2020}
\response{We should mention Wen \emph{et. al.} MDPI-Viruses 2018: \emph{Estimating Vaccine-Driven Selection in Seasonal Influenza}. }

\refb{Fig 1B is very hard to read. Maybe it should get more space (fig 1C could work with less space).}

\end{document}

### 2020-11-18 article.tex
SM ClusteringTraj - pfix w John's scores

>>>>

\subsection{Correcting for nested mutations} % (fold)
\label{sub:correcting_for_nested_mutations}

The analysis of the main text computes probabilities of fixation assuming that all trajectories are independent. 
However, it is well-known that mutations in infleunza viruses are nested: they appear on backgrounds that already carry other mutations. 
Since mutations appearing on the same genomes will jointly fix or disappear, many frequency trajectories are not independent but correlated. 
In order to compensate for potential biases due to this effect, we attempted to cluster trajectories based on similarity in their strain composition. 
Our aim is that two trajectories corresponding to mutations appearing mostly on the same genomes will be grouped in the same cluster. 
We then conduct the same analysis as in the main text on a set of \emph{effectively independent} trajectories constructed by taking one trajectory from each cluster.

In order to perform clustering, we define a distance between trajectories. 
A frequency trajectory $X$ is characterized by a series of frequency values and a time interval $T$. 
$f(t)$ for $t\in T$ corresponds to the frequency at which a given mutation $x$ appears in the population at date $t$, and we define $S(t)$ as the strains that carry mutation $x$ at date $t$. 
With this notation, $f(t)$ is the ratio of the number of elements in $S(t)$ to the total number of strains at date $t$. 
Let us now consider two frequency trajectories $X_1$ and $X_2$. \\
We define the distance $d(X_1,X_2)$ between these two trajectories based on the average similarity of the strains $S_1$ and $S_2$ that compose them: 
$$ d(X_1,X_2) = \frac{1}{\vert T_1\cap T_2\vert}\sum_{t\in T_1\cap T_2} \frac{\vert S_1(t)\cap S_2(t)\vert}{\vert S_1(t)\cup S_2(t)\vert}, $$ 
where $T_1\cap T_2$ is the time interval where both trajectories are active, and $\vert\cdot\vert$ denotes the number of elements of a set. 
The quantity summed corresponds to the Jaccard index between strains composing $X_1$ and $X_2$ at a given date. 
It is 1 if the two trajectories share exactly the same strains for this date, and 0 if they share no strain at all.  
This leads to the two following properties of $d$: 
\begin{itemize}
	\item if $d(X_1,X_2)=0$, then $X_1$ and $X_2$ represent the same frequency trajectory. The mutations $x_1$ and $x_2$ that they correpond to always appear on the same strains and are totally linked.
	\item if $d(X_1,X_2)=1$, then $X_1$ and $X_2$ can be considered completely independent. This can be the case if the two trajectories do not occur at the same dates, *i.e.* $\vert T_1\cap T_2\vert=0$, or if their respective mutations are never present on the same genomes.  
\end{itemize}

we attempt to reduce the potential statistical bias due to the nesting of trajectories by grouping them based on the above defined distance. 
Given a set of trajectories $\{X\}$, we perform a decomposition of $\{X\}$ into disjoint clusters $C_1(d^\star)\cup\ldots\cup C_n(d^\star)=\{X\}$ where $d^\star$ is an arbitrary threshold distance. 
Clusters are built in such a way that given two trajectories $X_i$ and $X_j$
$$ d(X_i,X_j) \leq d^\star \Rightarrow \exists k: X_i,X_j\in C_k $$
and
$$ X_i\in C_k \Rightarrow \exists j\in C_k: d(X_i,X_j) \leq d^\star. $$
These condition imply that the clusters formed are the minimal ones that guarantee that any two trajectories closer than the threshold distance $d^\star$ belong to the same cluster. 
The number of clusters $n$ depends on the chosen value for $d^\star$.

We compute clusters for different values of $d^\star$ for the case of the HA gene in A/H3N2. 
The top panel of figure \sref{fig:cluster_examples} shows the number cluster $n$ as a function of $d^\star$. 
In the $d^\star=0$ case, only trajectories that are exactly identical in terms of strain composition are clustered together. 
In this case, our clustering amounts to counting mutations that appear on exactly the same strains as one, reducing the number of effective trajectories from 800 to slightly less than 700. 
For higher values of $d^\star$, the number of clusters steadily goes down until it reaches 1 for $d^\star=1$, which is the maximum value of the distance $d(X_1,X_2)$. 
The sharp drop in $n$ for $d^\star=0.5$ is explained by the high number of very short (typically one time point) and low frequency trajectories that share one out of two strains.

\begin{figure}
	\centering
	\includegraphics[scale=0.5]{Rev_figures/clusters_examples.png}
	\caption{\textbf{Top}: Left-axis: number $n$ of clusters as a function of the threshold distance $d^\star$. The total number of trajectories is shown as a flat orange line. Right-axis: average size of clusters as a function of $d^\star$. \textbf{Bottom}: Examples of clusters for four values of $d^\star$. The four clusters displayed are the ones to which mutation HA1:33R belongs to. }
	\label{fig:cluster_examples}
\end{figure}

Since the choice of $d^\star$ is arbitrary and since no particular value can be chosen based on the number of clusters $n(d^\star)$, we decide to test our clustering strategy for five values, namely $d^\star\in\{0, 0.05, 0.1, 0.2, 0.49\}$.
The bottom panel of figure \sref{fig:cluster_examples} shows examples of a cluster for the four non-zero values of $d^\star$. 
The cluster displayed in each case is the one containing the mutation HA1:33R. 
As $d^\star$ increases, more and more unlike trajectories are grouped together. 
In the case $d^\star=0.49$, the cluster consists of 13 trajectories, 5 of which end up dying while the rest fix. 
Since such a high value of $d^\star$ results in grouping trajectories that do not have the same fate (fixation or death), we decide to exclude it from the rest of the analysis, resulting in four remaining values $d^\star\in\{0, 0.05, 0.1, 0.2\}$. \\

Once clustering is performed, we re-conduct the analysis of the main text on a set of effective trajectories. 
This set is constructed by taking one trajectory at random from each cluster. 
Effective trajectories are then considered independent from each other. 
The left panel of figure \sref{fig:pfix_v_freq_clusters} shows the fixation probability of trajectories as a function of their frequency for different values of $d^\star$, for the HA gene of A/H3N2.
The result obtained in panel A of figure \ref{fig:fix_loss_active} of the main text is also showed as a reference. 
For the three lower values of $d^\star$, results do not differ from the one obtained in the main text, even though the number of trajectories in each frequency bin has dropped as can be seen in the right panel of figure \sref{fig:pfix_v_freq_clusters}. 
This indicates that grouping together trajectories that share most of their strains, and are thus very correlated, does not modify the computed fixation probability in any way. 
For the higher value $d^\star=0.2$, fixation probability drops slightly across all frequency bins, suggesting that fixating trajectories tend to be grouped together more frequently. 
However, this drop remains of limited amplitude.

\begin{figure}
	\centering
	\includegraphics[scale=0.5]{Rev_figures/Pfix_v_freq_clustering_panel.png}
	\caption{\textbf{Left}: Fixation probability of trajectories as a function of probability for four values of $d^\star$. The reference curve is the same as in panel A of figure \ref{fig:fix_loss_active} for the HA gene. For readability, error bars are only displayed for the reference curve. \textbf{Bottom}: Number of trajectories in each frequency bin corresponding to the left panel.}
	\label{fig:pfix_v_freq_clusters}
\end{figure}

Overall, this analysis leads us to think that even though mutations in influenza may be nested, considering trajectories as independent does not result in strong statistical biases. 
Indeed, clustering similar trajectories together does significantly modify results presented in the main text.

% subsection correcting_for_nested_mutations (end)
